Capitalized Title Here

### 2021-03-04 CMPSpaper_writing/CMPSpaper_writing.Rmd
couple of small edits

In this paper we present an open-source implementation of the algorithm of the Congruent Matching Profile Segments (CMPS) method. @cmps developed the CMPS method for "objective comparison of striated tool marks" and demonstrated its usage in examples of comparing bullet signature correlations. Although @cmps introduced the algorithm of the CMPS method in their paper, \hh{they did not release an implementation of their method}. Thus, our effort here is to introduce the CMPS method to the `R` community and provide an open-source implementation of the algorithm for people to use and make further improvements. Our implementation is wrapped up into  `R` package  `CMPS`, which is available on GitHub at \url{https://github.com/willju-wangqian/CMPS}.

@@

An important task of firearm examination is \hh{to answer the question of whether two pieces of evidence come from the same source or whether a piece of evidence matches a sample obtained from a specific firearm. Here, in particular, we are interested } to determine whether two bullets were fired from the same gun barrel. 
\hh{Assessing the similarity between two bullets is based on a comparison of  striation marks acquired during the firing process as bullets are propelled through the barrel. }
\hh{Current state of art sees firearms examiners make an assessment of similarity based on a visual comparison, generally, using a comparison microscope. This practice has been criticized for its lack of objectivity and the associated problem in calculating an error rate [@pcast].}
<!--It is achievable because lands in a gun barrel create striation marks on the bullet during the firing process, and it is believed that the same gun barrel will create highly similar striation marks on the bullets it fired. Typically the striation marks are created by lands in a barrel, so the striation marks are also sometimes refered as land engravings, and the areas the marks occupy are called land engraved areas (LEA). However, in recent years people became more cautious about firearms identification, and criticism of its subjective methodology arose.--> A report published by @nrc states that "[m]uch forensic evidence-including, for example, bite marks and firearm and toolmark identification—is introduced in criminal trials without any meaningful scientific validation, determination of error rates, or reliability testing to explain the limits of the discipline." To overcome those critics and concerns, reseaerchers have been making an effort to build databases [@nistdb] and develop frameworks and algorithms that can bring an objective and quantitative assessment into the field. The bullet signatures extracted from the bullet are important in this process since they can well describe the striation marks on the bullet and can well be understood by algorithms.

@@

The datasets we worked with come from the James Hamby Consecutively Rifled Ruger Barrel Study [@brundage; @hamby; @Hamby:2019], and in particular, they are Hamby set 252 and Hamby set 44. For each Hamby set in the study, 35 bullets are fired by ten Ruger P-85 pistol barrels. 20 of the 35 bullets are known test bullets and 15 of them are unknown bullets. Each bullet has six land engraved areas (LEA) or land impressions, whose 3D topographical images were obtained and stored in the x3p format (XML 3-D Surface Profile). The x3p format provides a standard way of exchanging 2D and 3D profile data. It conforms to the ISO5436-2 standard [http://sourceforge.net/p/open-gps/mwiki/X3p/] and was adopted by the OpenFMC (Open Forensic Metrology Consortium), a group of firearm forensics researchers who contributes to the establishment of best practices of using metrology in forensic science. Hamby set 252 were scanned using a NanoFocus lens at 20x magnification with the scan resolution being 1.5625 \textmu m $\times$ 1.5625 \textmu m per pixel. Hamby set 44 was scanned \hh{at the Roy J Carver High-Resolution microscopy lab at Iowa State. These scans were acquired with a Sensofar Confocal Light Microscope at 20x magnification for a nominal resolution of 0.645 \textmu m $\times$ 0.645 \textmu per pixel}.  Both Hamby set 252 and Hamby set 44 are publicly available on the NIST Ballistics Database Project [@nistdb].

@@

The automated framework  used in this paper was proposed by @aoas. In order to obtain bullet signatures from the x3p files of bullet land impressions, we used `R` packages `x3ptools`[cite?] and `bulletxtrctr`[cite?]. `x3ptools` can be used to process x3p files. In the case of Hamby set 252, we need to convert the measurements of x3p files from meters to microns (`x3ptools::x3p_m_to_mum`) and rotate x3p files to make the long axis horizontal (`x3ptools::x3p_rotate`). Figure 1(a) shows a bullet land impression scan of a processed x3p file. The raised portion of the surface are groove impressions, and the vertical striation marks between the groove impressions are what we are interested in, called a land impression. The first step of obtaining the bullet signature is to extract a cross-section at a fixed height along the land impression. `bulletxtrctr::x3p_crosscut_optimize` can be used to determine an ideal height, and `bulletxtrctr::x3p_crosscut` can be used to acquire the crosscut data (`ccdata`). Figure 1(c) shows the typical pattern of the crosscut data, where "wedges" can be found on the left and right side and indicate the boundary between the land impression and the groove impression. `bulletxtrctr::cc_locate_grooves` is used to locate and chop these wedge areas so that only the curve of the land impression in the middle is kept. With the groove impressions excluded, we then use `bulletxtrctr::cc_get_signature` to smooth out the land impression curvature using LOESS (locally estimated scatter plot smoothing) [Cleveland 1979?] and obtain the bullet signature. Figure 1 illustrates this framework. Sometimes the algorithm in our framework cannot determine the ideal location of the crosscut or the grooves due to the deformation of a bullet, so we also developed a `R` Shiny App called `bulletinspectR` to identify and correct those errors. Note that NIST researchers used a different way of getting the bullet signatures in their paper [@cmps].

```{r, echo=FALSE, out.width="400px", fig.cap="image of a XXX gun barrel and a fired bullet"}
knitr::include_graphics("img/barrel_bullet.PNG", dpi = 100)
```

```{r, echo=FALSE, out.width="400px", fig.cap="(a) front view of a scanned land engraved area (LEA). The optimized crosscut location is indicated by the white line. (b) view of the cross-section of the land engraved area at the white line in (a). (c) the crosscut data plotted in 2D; blue vertical lines indicate the position of left and right grooves (d) the crosscut data after chopping the left and right grooves (e) the fitted curvature using LOESS (f) after removing the curvature from the crosscut data, the bullet signature is obtained"}
knitr::include_graphics("img/figure1_v2.PNG", dpi = 100)
```

>>>>

In this paper we present an open-source implementation of the algorithm of the Congruent Matching Profile Segments (CMPS) method. @cmps developed the CMPS method for "objective comparison of striated tool marks" and demonstrated its use in some examples of comparing bullet signature correlations. Although @cmps described the CMPS algorithm  in their paper, conceptually, \hh{the authors did not release an implementation of their method}. Thus, our effort here is to introduce the CMPS method to the `R` community and provide an open-source implementation of the algorithm for people to use, review, and further improvements. Our implementation is wrapped up into  `R` package  `CMPS`, available on GitHub at \url{https://github.com/willju-wangqian/CMPS}.

@@

An important task of firearm examination is \hh{to answer the question of whether two pieces of evidence come from the same source or whether a piece of evidence matches a sample obtained from a specific firearm. Here, in particular, we are interested } to determine whether two bullets were fired from the same gun barrel. 
\hh{Assessing the similarity between two bullets is based on a comparison of  striation marks acquired during the firing process as bullets are propelled through the barrel. }
\hh{Current state of art sees firearms examiners make an assessment of similarity based on a visual comparison, generally, using a comparison microscope. This practice has been criticized for its lack of objectivity and the associated problem in calculating an error rate [@pcast].}
<!--It is achievable because lands in a gun barrel create striation marks on the bullet during the firing process, and it is believed that the same gun barrel will create highly similar striation marks on the bullets it fired. Typically the striation marks are created by lands in a barrel, so the striation marks are also sometimes refered as land engravings, and the areas the marks occupy are called land engraved areas (LEA). However, in recent years people became more cautious about firearms identification, and criticism of its subjective methodology arose.--> A report published by @nrc states that "[m]uch forensic evidence-including, for example, bite marks and firearm and toolmark identification—is introduced in criminal trials without any meaningful scientific validation, determination of error rates, or reliability testing to explain the limits of the discipline." To overcome those criticisms and concerns, researchers have been making an effort to build databases [@nistdb] and develop frameworks and algorithms that  bring an objective and quantitative assessment into the field \hh{XXX cite some of these efforts - make sure to say that it is not an exhaustive list}. 
\hh{XXX The bullet signatures weren't mentioned before - make sure to include that there is a whole section that discusses how bullet signatures are extracted from scans.}
The bullet signatures extracted from the bullet are important in this process since they can well describe the striation marks on the bullet and can well be understood by algorithms.

@@

The datasets we worked with come from the James Hamby Consecutively Rifled Ruger Barrel Study [@brundage; @hamby; @Hamby:2019],  in particular, Hamby set 252 and Hamby set 44. 
For each Hamby set, a total of 35 bullets is fired from ten \hh{consecutively manufactured} Ruger P-85 pistol barrels. \hh{Two bullets are fired from each barrel, making up a set of 20 reference bullets. An additional 15 bullets are fired from these ten barrels in a fashion unknown to the study participant. The aim of the Hamby Study was to have firearms examiners identify which barrel each of the 15 questioned bullets was fired from. The Ruger P-85 barrels are traditionally rifled barrels with six grooves and lands \hh{as shown in \autoref{fig:bullet}}. During the firing process, grooves and lands are engraved on a bullet. Firearms examiners use striation marks on land engraved areas (LEAs) for their visual comparison. For algorithmic purposes, }
 3D topographical images \hh{land engraved areas} were obtained and stored in  x3p format (XML 3-D Surface Profile). The x3p format provides a standard way of exchanging 2D and 3D profile data. It conforms to the ISO5436-2 standard [http://sourceforge.net/p/open-gps/mwiki/X3p/]  adopted by the OpenFMC (Open Forensic Metrology Consortium), a group of firearm forensics researchers who contributes to the establishment of best practices of using metrology in forensic science. Hamby set 252 were scanned using a NanoFocus lens at 20x magnification with the scan resolution being 1.5625 \textmu m $\times$ 1.5625 \textmu m per pixel. Hamby set 44 was scanned \hh{at the Roy J Carver High-Resolution microscopy lab at Iowa State. These scans were acquired with a Sensofar Confocal Light Microscope at 20x magnification for a nominal resolution of 0.645 \textmu m $\times$ 0.645 \textmu per pixel}.  Both Hamby set 252 and Hamby set 44 are publicly available on the NIST Ballistics Database Project [@nistdb].

@@

The automated framework  used in this paper was proposed by @aoas. In order to obtain bullet signatures from the x3p files of bullet land impressions, we use the `R` packages `x3ptools` [@x3ptools] and `bulletxtrctr` [@bulletxtrctr]. `x3ptools` \hh{is a package to read, write, and generally, process } x3p files. 
<!--In the case of Hamby set 252, we need to convert the measurements of x3p files from meters to microns (`x3ptools::x3p_m_to_mum`) and rotate x3p files to make the long axis horizontal (`x3ptools::x3p_rotate`).--> 
\hh{it's land **engraved** areas, not impressions (that was wrong in Eric's first paper). Impressions are marks from impacts, such as the hit of a hammer or the hit from the breech face. Engravings are scratch marks.}
\autoref{fig:process}(a) shows a \hh{rendering of a 3d scan of a } bullet land  engraved area. The raised portion of the surface \hh{on the left and right of the scan are parts of the adjacent groove engraved areas (GEAs)}, \hh{the middle area shows a land-engraved area (LEA) with well expressed} striation marks. The first step of obtaining the bullet signature is to extract a cross-section at a fixed height along the land impression. `bulletxtrctr::x3p_crosscut_optimize` can be used to determine an ideal height, and `bulletxtrctr::x3p_crosscut` can be used to acquire the crosscut data (`ccdata`). \autoref{fig:process}(c) shows the typical pattern of the crosscut data, where "wedges" can be found on the left and right side and indicate the boundary between the land impression and the groove impression. `bulletxtrctr::cc_locate_grooves` is used to locate and chop these wedge areas so that only the curve of the land impression in the middle is kept. With the groove impressions excluded, we then use `bulletxtrctr::cc_get_signature` to smooth out the land impression curvature using LOESS (locally estimated scatter plot smoothing) [@loess] and obtain the bullet signature. \autoref{fig:process} illustrates this framework. Sometimes the algorithm in our framework cannot determine the ideal location of the crosscut or the grooves due to the deformation of a bullet, so we also developed a `R` Shiny App called `bulletinspectR` to identify and correct those errors. Note that NIST researchers used a different way of getting the bullet signatures in their paper [@cmps].

```{r bullet, echo=FALSE, out.width=".8\\textwidth", fig.cap="image of a XXX gun barrel and a fired bullet - \\hh{xxx my version of the image still has a groove engraved area marked as a land}"}
knitr::include_graphics("img/barrel_bullet.PNG", dpi = 100)
```

```{r process, echo=FALSE, out.width=".9\\textwidth", fig.cap="(a) front view of a scanned land engraved area (LEA). The optimized crosscut location is indicated by the white line. (b) view of the cross-section of the land engraved area at the white line in (a). (c) the crosscut data plotted in 2D; blue vertical lines indicate the position of left and right grooves (d) the crosscut data after chopping the left and right grooves (e) the fitted curvature using LOESS (f) after removing the curvature from the crosscut data, the bullet signature is obtained"}
knitr::include_graphics("img/figure1_v2.PNG", dpi = 100)
```
